---
title: "Activity Recognition"
author: "Emel Aktas"
date: "26 July 2015"
output: html_document
---


Introduction
------------

The aim of this project is to recognise activities of people using data recorded by wearable devices such as belts, gloves, arm bands as well as dumbbells equipped with a sensor. Participants are asked to perform Unilateral Dumbbell Biceps Curl in five different fashions; hence, there are five classes of activities to be recognised by the machine learning algorithm. The first activity is correct implementation of lifting whereas the remaining four represent common mistakes as follows:

- Class A: Exactly according to the specification 
- Class B: Throwing the elbows to the front 
- Class C: Lifting the dumbbell only halfway
- Class D: Lowering the dumbbell only halfway 
- Class E: Throwing the hips to the front

To predict the activity class, a machine learning algorithm is built and then cross-validated. Subsequently, the out-of-sample error is estimated. This report primarily presents these three components of the project and also explains the rationale behind the choices made in this process. The report ends with brief conclusions.

Model Building
---------------
The aim of the machine learning model is to predict the activity quality classes A-E as described above from sensor readings. 

```{r}
# Read training data
training <- read.csv("pml-training.csv", na.strings=c("NA",""), header=T)

# Read problem sets
predictionset <- read.csv("pml-testing.csv", na.strings=c("NA",""), header=T)
# str(training) # check structure - commented out to save space
dim(training)
# summary(training) # see the summary of variables - commented out to save space

# Load necessary packages
library(AppliedPredictiveModeling)
library(caret)
library(rattle)
library(rpart.plot)
library(randomForest)

# Count how many NAs in each column to exclude those columns from the machine learning algorithm
NAs <- function(x) {
    as.vector(apply(x, 2, function(x) length(which(is.na(x)))))
}

# Count NAs in each column
columnNAcount <- NAs(training)
columnNAcount

# Find the variables that do not have NAs 
# Initialise variable to keep column names
newdatavars <- c()

# Develop the variable
for (i in 1:160){
  if(columnNAcount[i] == 0 ){
    newdatavars <- c(newdatavars, colnames(training)[i])
  }
}

newdatavars


# choose smaller data set
newtraining <- training[newdatavars]
dim(newtraining)

# Covariates with no variability
nsv <- nearZeroVar(newtraining, saveMetrics=TRUE)
nsv

# A quick look suggests new_window is an NZV. Delete new_window
newtraining <- newtraining[, -which(names(newtraining) %in% c("new_window"))]
str(newtraining)

# Remove columns 1-6 which will not be needed for prediction
newtraining <- newtraining[-c(1:6)]

# Split data into training and testing data
inTrain <- createDataPartition(newtraining$classe, p = .75, list = F)
trainingdata <- newtraining[inTrain,]
testingdata <- newtraining[-inTrain,]

```

Cross-validation
----------------
```{r}
# To use three repeats of 10–fold cross–validation
cvCtrl <- trainControl(method = "repeatedcv", repeats = 3)

# Build the machine learning model classification tree
modFit1 <- train(classe ~ ., data = trainingdata, preProcess=c("center", "scale"), method = "rpart", trControl = cvCtrl)

# Plot the tree
fancyRpartPlot(modFit1$finalModel)


# Build the machine learning model random forest
modFit2 <- randomForest(classe ~ . , data = trainingdata)


```


Expected out-of-sample error
----------------------------
```{r}
# Predict with the CART
prediction1 <- predict(modFit1, newdata = testingdata)
confusionMatrix(prediction1, testingdata$classe)

# Predict with the Random Forest
prediction2 <- predict(modFit2, newdata = testingdata)
confusionMatrix(prediction2, testingdata$classe)

```




Rationale and Conclusion
------------------------
The prediction accuracy is much better with the Random Forests; hence I submitted the prediction from modFit2.

```{r}
# Predict test set with the Random Forest
predict(modFit2, newdata = predictionset)
```




References
----------

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises#ixzz3geVCgjEB

